Canonical rings of $\bq$-divisors on Minimal Rational Surfaces

### 2015-08-04 surfaces-write-up/surface-write-up.tex
started indexing at 0 and other responses

\usepackage{amsmath}
\usepackage{amsxtra}
\usepackage{amscd}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{eucal}
\usepackage[all]{xy}
\usepackage{graphicx}
\usepackage{tikz-cd}
\usepackage{mathrsfs}
\usepackage{subfiles}
%\usepackage{mathpazo} not a huge fan
\usepackage{euler}
\usepackage{hyperref}
\usepackage{color}
\usepackage{longtable}
\usepackage{float}
\usepackage{caption}
\usepackage{thmtools}
\usepackage{thm-restate}

@@

In Section \ref{sec:lemmas}, we discuss background knowledge and present some useful lemmas.  In Section \ref{sec:proj} we focus on projective spaces: $\mathbb{P}^n$ and prove the following two results:
\begin{restatable}{thm}{restateeffective}
\label{thm:proj-effective-intro}
Let $D = \sum_{i = 1}^{n} \alpha_i C_i \in \bz \otimes_\bz \di \bp^m$, with $\alpha_i =
\frac{p_i}{q_i} \in \bq$ and each $C_i$ a hypersurface of
$\bp^m$, be an effective divisor on $\bp^m$.

@@

\begin{restatable}{thm}{restateproj}
\label{thm:proj-generators-relations}
Let $D = \sum_{i=1}^n \alpha_i D_i \in \bq \otimes_\bz \di \bp^m$. Let
$f_i$ be homogeneous polynomials in $x_0, \ldots, x_m$ such that
$D_i = V(f_i)$ for all $i$; suppose $\deg(f_i) = a_i$. 
Then $R_D$ is generated in degrees at most 
\[
	\sum_{i=1}^n \ell_i a_i
\]

\noindent
with relations generated in degrees at most
\[
	\max \left(2 \sum_{i=1}^n \ell_i a_i, \frac{\max_{1\le i \le n}
	(	\bida_i)}{\deg(D)} + \sum_{i=1}^n \ell_i a_i \right).
\]
\end{restatable}

@@

@@

\begin{prop}
\label{prop:pm-span-and-basis}
The functions $u^d \cdot \prod_{i=1}^n f_i^{c_i}$ satisfying \todo{Peter: Why do we include this $u$?  IN the later sections, I don't use this notation and we should standardize.}
\begin{align}
\label{align:pm-span}
\sum_{i=1}^{n} c_i \cdot \bida_i = 0 && \text{ and } &&c_i \geq \lfloor \alpha_i d\rfloor	
\end{align}

\noindent
form a spanning set for $H^0(\bp^m, dD)$ over $k$. Furthermore, functions 
of the form $u^d \cdot \prod_{i=1}^n f_i^{c_i}$ satisfying
\begin{align}
\label{align:pm-basis}
c_i = -\lfloor d\alpha_i \rfloor && \text{ for } && i > m
\end{align}\todo{Peter: Should be for $i> m+1$ right?  Looking below I see an $f_0$, where as above $f_i$ ranges 1 to $n$.  We should choose whether or not 0 is included in the indexing and fix the section to match that convention.}

@@

In this subsection, we restrict attention to the case of effective
fractional divisors $D \in \bq \otimes \di \bp^m$. We completely
characterize the canonical rings when $D$ consists of one
hypersurface and give strong \todo{Peter: I'm opposed to the inclusion of the word strong, as it ambiguous and subjective.  I think we should either replace it with tight, and then explicitly say later how it is tight, or just get rid of the adjective.} bounds on generator and relation
degrees for general effective divisors $D$.

\begin{convention}
Let $[x_0, \ldots, x_m]$ be coordinates for $\bp^m$. Let
$\vec{v} = (v_0, \ldots, v_m) \in \bq^{m + 1}$. \todo{Peter: do we want to allow $\vec{v}\in \mathbb{Q}^{m+1}$ rather than $\mathbb{Z}^{m+1}$?  I'm not sure if the arguments actually need the generality (and if so, it is a little worrisome since fractional powers aren't clearly defined algebraically.}  Then write
\[
	x^{\vec{v}} := \prod_{i = 0}^{m} x_i^{v_i}
\]
\end{convention}

@@

Let $D = \sum_{i=1}^n \alpha_i D_i \in \bq \otimes \di \bp^m$ and let $f_i$ be
homogeneous polynomials in $x_0, \ldots, x_m$ such that $D_i = V(f_i)$
$\deg(f_i) = a_i$ for all $i$.

For the remainder of this section, we shall make the additional assumption that 
\begin{align}
\label{eqn:pm-basis-assumption}
	f_1, \ldots, f_{m+1} \text{ form a } \bk \text{ basis of all degree 1 functions.}
\end{align}

@@

This example further generalizes to the following situation:
Suppose $D = \sum_{i=1}^{n} \frac{\num_i}{\den_i}D_i \in \di \bp^m
\otimes \bq,$ where $\deg D_i = \bida_i$ and $\deg D = \frac{1}{\lcm
_{1 \leq i \leq n}(\den_i \cdot \bida_i)}$. Then, if $D_i = V(f_i)$
where all $f_i$ can be written as a polynomial function in $x_0,
\ldots, x_{m-1},$ we have $R_D$ always has a generator in degree
$\lcm(\den_i \cdot \bida_i)$.
\end{example}

\begin{rem}\todo{Peter: maybe move this out of a remark and just say it with the previous example?  As it stands, the previous example is a little unclear as to what its purpose is...}
\label{rem:ghost-motivation}
Example ~\ref{eg:hyperplane} shows that the naive generalization of
~\cite[Theorem 8]{dorney:canonical} of generation in degree at most
$\sum_{i=1}^{n}\ell_i$ cannot possibly hold. The reason for this is
that the divisors may ``be contained in a hyperplane,'' i.e.~ be
expressible as functions in $m$ of the $m+1$ variables on $\bp^m$.
This problem can easily be circumvented by adding in ``ghost
points.'' That is, we may add divisors of the form $0 \cdot H_i$ to
$D$, and reorder so that if $D = \sum_{i=1}^{n}\alpha_i V(f_i)$,
then $f_0, \ldots, f_m$ cut out linearly independent
hyperplanes in $\bp^m$.

@@

\begin{defn}\label{def:ell_i}
Let $D = \sum_{i=1}^n \alpha_i D_i$ where $\alpha_i = \frac{c_i}{k_i}$.
Then notate
\[\ell_i = \lcm_{j\ne i}(k_i).\]
\end{defn}

\begin{lem} \label{lem:proj-generators}
$R_D$ is generated in degrees at most $\sum_{i=1}^n \ell_i a_i.$

@@

We note that $\Sigma$ has extremal rays given by the lattice points 
\begin{equation}\label{defn:e-i-proj}
	e_i = \left(\ell_i \bida_i, - \alpha_1 \ell_i \bida_i, \ldots
-\alpha_{i-1} \ell_i \bida_i, \ell_i \sum_{j\ne i} \alpha_j \bida_j,
-\alpha_{i+1} \ell_i \bida_i, \ldots, -\alpha_n, \ell_i \bida_i \right)
\end{equation}
for each $i\in \{1, \ldots n\}$.
Therefore, applying Proposition ~\ref{prop:cone-generation}, we see $R_D$ is generated in degrees less than
\[
	\sum_{i=1}^n \ell_i a_i.
\]
\end{proof}

@@

Let $x_1, \ldots x_r$ be the generators in degrees at most $\sum_{i=1}^n \ell_i
\bida_i$ (given by Lemma ~\ref{lem:proj-generators}), and let $\phi$ be the
surjection $k[x_1, \ldots x_r] \to R_D$.  We can factor $\phi$ through the
semigroup ring 
\[
	\bk[\Sigma] =  \langle u^d z_1^{c_1} \cdots z_n^{c_n} : c_i \in
\mathbb{Z}, \; c_i \geq -d \alpha_i, \mbox{ and }\sum_{i=1}^{n} \bida_i c_i =
\sum_{i=1}^{n} \bidb_i c_i \rangle. 
\]
by
\[
\begin{tikzcd}
\bk[x_1,\ldots, x_r] \ar {r}{\chi} & \bk[\Sigma] \ar {r}{\psi} & R_D \\
x_i \ar r & u^{d_i}z_1^{c_{i1}} \cdots z_n^{c_{in}} \ar r & u^{d_i}f_1^{c_{i1}} \cdots f_n^{c_{in}}.
\end{tikzcd}
\]

@@

\begin{proof}
We first claim there exist $\beta_1, \ldots, \beta_n \in k[\Sigma]$
such that $\ker \psi$ is generated by
\begin{equation}
\label{eqn:relations-psi-proj}
	(z_i - \beta_i)\prod_{j=1}^n {z_j}^{c_{j}}
\end{equation}

\noindent
for all $d \in \mathbb{N}$ and $c_i \ge -\alpha_i d$ satisfying
$\bida_i + \sum_{j = 1}^n \bida_j c_j = 0$.

We define the $\beta_i$ above so that $\beta_i$ is the expansion of
$f_i$ when viewed as a polynomial in $k[f_1,\ldots, f_{m + 1}]$.
Furthermore, the relations given in Equation
~\eqref{eqn:proj-relation-degree} generate all relations, since they
allow us to reduce any $ \prod_{j = 1}^n f_j^{r_i}$ to a canonical
form, with $r_i = -\lfloor  d \alpha_i\rfloor$ whenever $i  > m + 1$.

For the remainder of the proof, fix $i \in \{1,\ldots, n\}$. To
complete the proof, it suffices to bound the degree of generation
of the relations of the form $(z_i - \beta_i) \prod_{j = 1}^n
z_j^{c_j}$, by Equation ~\eqref{eqn:proj-relation-degree}. We first
associate $(z_i - \beta_i)\prod_{j=1}^n z_j^{c_j}$ with $(d, c_1,
\ldots, c_n) \in \Sigma,$ as defined in Equation
~\eqref{eqn:Sigma-def}. Let $\Sigma_i \subseteq \mathbb{Z}^{n + 1}$
be the set of points of the form $(d, r_1, \ldots, r_n)$ satisfying
$r_j \ge -d \alpha_j$ for all $j$ and $\sum_{j=1}^n r_j a_j = -a_i$.
Let
\begin{align*}
	\delta_i = \left(\frac{a_i}{\deg(D)}, -\frac{\alpha_1 a_1}{\deg(D)},
	\ldots, - \frac{\alpha_n a_n}{\deg(D)} \right).
\end{align*}

\noindent
Then we see $\Sigma_i \subseteq \delta_i + \newspan_{\bq_{\geq 0}}
(e_1, \ldots, e_n)$ with $e_i$ as defined in Equation
~\ref{defn:e-i-proj}. Therefore, we can write any element of
$\Sigma_i$ as
\[
	\delta_i + \sum_{j=1}^n r_j e_j
\]
where $r_j \in \mathbb{R}$ for each $j$.

Whenever some $r_j \ge 1$, we can generate the relation $(z_i -
\beta_i)\prod_{j=1}^n z_j^{c_j} = \epsilon_j \cdot h,$ where
$\epsilon_j$ was defined in Equation ~\ref{eqn:epsilon-def-proj} and
$h \in \Sigma_i$. Therefore, $\ker \psi$ is generated in degrees
less than
\[
	\frac{\bida_i}{\deg(D)} + \sum_{i=1}^n \ell_i a_i.
\]
\end{proof}

@@

By ~\ref{lem:bound-ker-chi}, $\ker \chi$ is generated in degrees 
at most $2\sum_{i=1}^n \ell_i a_i$ and by Lemma
~\ref{lem:proj-relations-psi}, $\ker \psi,$ is generated in
degrees up to $\frac{\bida_i}{\deg(D)} + \sum_{i=1}^n \ell_i a_i$. 
Consequently, Lemma ~\ref{lem:composite-map} implies that $\ker \phi$
is generated in degrees less than
\[
	\max \left(2 \sum_{i=1}^n \ell_i a_i, \frac{\max_{1\le i \le n}
	(\bida_i)}{\deg(D)} + \sum_{i=1}^n \ell_i a_i \right).
\]
\end{proof}

>>>>

\usepackage{amsmath}
\usepackage{amsxtra}
\usepackage{amscd}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{eucal}
\usepackage[all]{xy}
\usepackage{graphicx}
\usepackage{tikz-cd}
\usepackage{mathrsfs}
\usepackage{subfiles}
\usepackage{euler}
\usepackage{hyperref}
\usepackage{color}
\usepackage{longtable}
\usepackage{float}
\usepackage{caption}
\usepackage{thmtools}
\usepackage{thm-restate}

@@

In Section \ref{sec:lemmas}, we discuss background knowledge and present some useful lemmas.  In Section \ref{sec:proj} we focus on projective spaces: $\mathbb{P}^n$ and prove the following two results:
\begin{restatable}{thm}{restateeffective}
\label{thm:proj-effective-intro}
Let $D = \sum_{i = 0}^{n} \alpha_i C_i \in \bz \otimes_\bz \di \bp^m$, with $\alpha_i =
\frac{p_i}{q_i} \in \bq$ and each $C_i$ a hypersurface of
$\bp^m$, be an effective divisor on $\bp^m$.

@@

\begin{restatable}{thm}{restateproj}
\label{thm:proj-generators-relations}
Let $D = \sum_{i=0}^n \alpha_i D_i \in \bq \otimes_\bz \di \bp^m$. Let
$f_i$ be homogeneous polynomials in $x_0, \ldots, x_m$ such that
$D_i = V(f_i)$ for all $i$; suppose $\deg(f_i) = a_i$. 
Then $R_D$ is generated in degrees at most 
\[
	\sum_{i=0}^n \ell_i a_i
\]

\noindent
with relations generated in degrees at most
\[
	\max \left(2 \sum_{i=0}^n \ell_i a_i, \frac{\max_{1\le i \le n}
	(	\bida_i)}{\deg(D)} + \sum_{i=0}^n \ell_i a_i \right).
\]
\end{restatable}

@@

\todo{Aaron: weren't we going to take this section out and move it into the later sections?}

@@

\begin{prop}
\label{prop:pm-span-and-basis}
The functions $u^d \cdot \prod_{i=1}^n f_i^{c_i}$ satisfying \todo{Peter: Why
do we include this $u$?  IN the later sections, I don't use this notation and
we should standardize.

Aaron: Isn't it necessary to distinguish the functions? For instance, we could consider $f_0$ either as a function in degree 1 or 2 for $D = \frac{1}{2}x_0$, and we need
$u^d$ to distinguish which degree it lives in.
I think we will need to go through and include the $d$ throughout.
Do you think there's a way to avoid including the $d$?
}
\begin{align}
\label{align:pm-span}
\sum_{i=0}^{n} c_i \cdot \bida_i = 0 && \text{ and } &&c_i \geq \lfloor \alpha_i d\rfloor	
\end{align}

\noindent
form a spanning set for $H^0(\bp^m, dD)$ over $k$. Furthermore, functions 
of the form $u^d \cdot \prod_{i=0}^n f_i^{c_i}$ satisfying
\begin{align}
\label{align:pm-basis}
c_i = -\lfloor d\alpha_i \rfloor && \text{ for } && i > m
\end{align}\todo{Peter: Should be for $i> m+1$ right?  Looking below I see an $f_0$, where as above $f_i$ ranges 1 to $n$.  We should choose whether or not 0 is included in the indexing and fix the section to match that convention.

Aaron: I tried to fix it at $0$ everywhere. If probably missed a few places, so keep your eyes peeled.
}

@@

In this subsection, we restrict attention to the case of effective
fractional divisors $D \in \bq \otimes \di \bp^m$. We completely
characterize the canonical rings when $D$ consists of one
hypersurface and give tight bounds on generator and relation
degrees for general effective divisors $D$.
\todo{Aaron: make a remark as to why these bounds are tight}
\begin{convention}
Let $[x_0, \ldots, x_m]$ be coordinates for $\bp^m$. Let
$\vec{v} = (v_0, \ldots, v_m) \in \bq^{m + 1}$. \todo{Peter: do we want to allow $\vec{v}\in \mathbb{Q}^{m+1}$ rather than $\mathbb{Z}^{m+1}$?  I'm not sure if the arguments actually need the generality (and if so, it is a little worrisome since fractional powers aren't clearly defined algebraically.

Aaron: No, these vectors should just lie in $\bz$. They're just used
to given an ordering on monomials.
}  Then write
\[
	x^{\vec{v}} := \prod_{i = 0}^{m} x_i^{v_i}
\]
\end{convention}

@@

Let $D = \sum_{i=0}^n \alpha_i D_i \in \bq \otimes \di \bp^m$ and let $f_i$ be
homogeneous polynomials in $x_0, \ldots, x_m$ such that $D_i = V(f_i)$
$\deg(f_i) = a_i$ for all $i$.

For the remainder of this section, we shall make the additional assumption that 
\begin{align}
\label{eqn:pm-basis-assumption}
	f_0, \ldots, f_{m} \text{ form a } \bk \text{ basis of all degree 1 functions.}
\end{align}

@@

This example further generalizes to the following situation:
Suppose $D = \sum_{i=0}^{n} \frac{\num_i}{\den_i}D_i \in \di \bp^m
\otimes \bq,$ where $\deg D_i = \bida_i$ and $\deg D = \frac{1}{\lcm
_{1 \leq i \leq n}(\den_i \cdot \bida_i)}$. Then, if $D_i = V(f_i)$
where all $f_i$ can be written as a polynomial function in $x_0,
\ldots, x_{m-1},$ we have $R_D$ always has a generator in degree
$\lcm(\den_i \cdot \bida_i)$.
\end{example}

\begin{rem}\todo{Peter: maybe move this out of a remark and just say it with the previous example?  As it stands, the previous example is a little unclear as to what its purpose is...}
\label{rem:ghost-motivation}
Example ~\ref{eg:hyperplane} shows that the naive generalization of
~\cite[Theorem 8]{dorney:canonical} of generation in degree at most
$\sum_{i=0}^{n}\ell_i$ cannot possibly hold. The reason for this is
that the divisors may ``be contained in a hyperplane,'' i.e.~ be
expressible as functions in $m$ of the $m+1$ variables on $\bp^m$.
This problem can easily be circumvented by adding in ``ghost
points.'' That is, we may add divisors of the form $0 \cdot H_i$ to
$D$, and reorder so that if $D = \sum_{i=0}^{n}\alpha_i V(f_i)$,
then $f_0, \ldots, f_m$ cut out linearly independent
hyperplanes in $\bp^m$.

@@

\begin{defn}\label{def:ell_i}
Let $D = \sum_{i=0}^n \alpha_i D_i$ where $\alpha_i = \frac{c_i}{k_i}$.
Then notate
\[\ell_i = \lcm_{j\ne i}(k_i).\]
\end{defn}

\begin{lem} \label{lem:proj-generators}
$R_D$ is generated in degrees at most $\sum_{i=0}^n \ell_i a_i.$

@@

We note that $\Sigma$ has extremal rays given by the lattice points 
\begin{equation}\label{defn:e-i-proj}
	e_i = \left(\ell_i \bida_i, - \alpha_1 \ell_i \bida_i, \ldots
-\alpha_{i-1} \ell_i \bida_i, \ell_i \sum_{j\ne i} \alpha_j \bida_j,
-\alpha_{i+1} \ell_i \bida_i, \ldots, -\alpha_n, \ell_i \bida_i \right)
\end{equation}
for each $i\in \{1, \ldots n\}$.
Therefore, applying Proposition ~\ref{prop:cone-generation}, we see $R_D$ is generated in degrees less than
\[
	\sum_{i=0}^n \ell_i a_i.
\]
\end{proof}

@@

Let $x_1, \ldots x_r$ be the generators in degrees at most $\sum_{i=1}^n \ell_i
\bida_i$ (given by Lemma ~\ref{lem:proj-generators}), and let $\phi$ be the
surjection $k[x_1, \ldots x_r] \to R_D$.  We can factor $\phi$ through the
semigroup ring 
\[
	\bk[\Sigma] =  \langle u^d z_0^{c_0} \cdots z_n^{c_n} : c_i \in
\mathbb{Z}, \; c_i \geq -d \alpha_i, \mbox{ and }\sum_{i=0}^{n} \bida_i c_i =
\sum_{i=0}^{n} \bidb_i c_i \rangle. 
\]
by
\[
\begin{tikzcd}
\bk[x_1,\ldots, x_r] \ar {r}{\chi} & \bk[\Sigma] \ar {r}{\psi} & R_D \\
x_i \ar r & u^{d_i}z_0^{c_{i0}} \cdots z_n^{c_{in}} \ar r & u^{d_i}f_1^{c_{i1}} \cdots f_n^{c_{in}}.
\end{tikzcd}
\]

@@

\begin{proof}
We first claim there exist $\beta_1, \ldots, \beta_n \in k[\Sigma]$
such that $\ker \psi$ is generated by
\begin{equation}
\label{eqn:relations-psi-proj}
	(z_i - \beta_i)\prod_{j=0}^n {z_j}^{c_{j}}
\end{equation}

\noindent
for all $d \in \mathbb{N}$ and $c_i \ge -\alpha_i d$ satisfying
$\bida_i + \sum_{j = 0}^n \bida_j c_j = 0$.

We define the $\beta_i$ above so that $\beta_i$ is the expansion of
$f_i$ when viewed as a polynomial in $k[f_0,\ldots, f_{m}]$.
Furthermore, the relations given in Equation
~\eqref{eqn:proj-relation-degree} generate all relations, since they
allow us to reduce any $ \prod_{j = 0}^n f_j^{r_i}$ to a canonical
form, with $r_i = -\lfloor  d \alpha_i\rfloor$ whenever $i  > m$.

For the remainder of the proof, fix $i \in \{1,\ldots, n\}$. To
complete the proof, it suffices to bound the degree of generation
of the relations of the form $(z_i - \beta_i) \prod_{j = 1}^n
z_j^{c_j}$, by Equation ~\eqref{eqn:proj-relation-degree}. We first
associate $(z_i - \beta_i)\prod_{j=0}^n z_j^{c_j}$ with $(d, c_1,
\ldots, c_n) \in \Sigma,$ as defined in Equation
~\eqref{eqn:Sigma-def}. Let $\Sigma_i \subseteq \mathbb{Z}^{n + 2}$
be the set of points of the form $(d, r_1, \ldots, r_n)$ satisfying
$r_j \ge -d \alpha_j$ for all $j$ and $\sum_{j=0}^n r_j a_j = -a_i$.
Let
\begin{align*}
	\delta_i = \left(\frac{a_i}{\deg(D)}, -\frac{\alpha_0 a_0}{\deg(D)},
	\ldots, - \frac{\alpha_n a_n}{\deg(D)} \right).
\end{align*}

\noindent
Then we see $\Sigma_i \subseteq \delta_i + \newspan_{\bq_{\geq 0}}
(e_0, \ldots, e_n)$ with $e_i$ as defined in Equation
~\ref{defn:e-i-proj}. Therefore, we can write any element of
$\Sigma_i$ as
\[
	\delta_i + \sum_{j=0}^n r_j e_j
\]
where $r_j \in \mathbb{R}$ for each $j$.

Whenever some $r_j \ge 1$, we can generate the relation $(z_i -
\beta_i)\prod_{j=1}^n z_j^{c_j} = \epsilon_j \cdot h,$ where
$\epsilon_j$ was defined in Equation ~\ref{eqn:epsilon-def-proj} and
$h \in \Sigma_i$. Therefore, $\ker \psi$ is generated in degrees
less than
\[
	\frac{\bida_i}{\deg(D)} + \sum_{i=0}^n \ell_i a_i.
\]
\end{proof}

@@

By ~\ref{lem:bound-ker-chi}, $\ker \chi$ is generated in degrees 
at most $2\sum_{i=0}^n \ell_i a_i$ and by Lemma
~\ref{lem:proj-relations-psi}, $\ker \psi,$ is generated in
degrees up to $\frac{\bida_i}{\deg(D)} + \sum_{i=0}^n \ell_i a_i$. 
Consequently, Lemma ~\ref{lem:composite-map} implies that $\ker \phi$
is generated in degrees less than
\[
	\max \left(2 \sum_{i=0}^n \ell_i a_i, \frac{\max_{1\le i \le n}
	(\bida_i)}{\deg(D)} + \sum_{i=0}^n \ell_i a_i \right).
\]
\end{proof}
